---
title: "Binomial Ordering Preferences — Ablation Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(lme4)
library(lmerTest)

theme_set(theme_bw(base_size = 12))

# ── Paths ──────────────────────────────────────────────────────────────────
DATA_DIR    <- "../Data/model_pref_results"
BIGRAM_CSV  <- "../Data/bigram_stats.csv"
FREQ_CSV    <- "../Data/frequency_log.csv"
```

## Load & join data

```{r load}
# Load all per-checkpoint CSVs and stack them
pref_raw <- list.files(DATA_DIR, pattern = "\\.csv$", full.names = TRUE) |>
  map(read_csv, show_col_types = FALSE) |>
  bind_rows()

# Parse model size from model name
pref_raw <- pref_raw |>
  mutate(
    model_size = case_when(
      str_detect(model, "1\\.3b") ~ "1.3B",
      str_detect(model, "350m")   ~ "350M",
      str_detect(model, "125m")   ~ "125M",
      TRUE                        ~ "unknown"
    ),
    model_size = factor(model_size, levels = c("125M", "350M", "1.3B"))
  )

# Load bigram stats (one row per binomial)
bigrams <- read_csv(BIGRAM_CSV, show_col_types = FALSE) |>
  mutate(bigram_log_ratio = log(bigram_alpha) - log(bigram_nonalpha)) |>
  select(binom_alpha, bigram_alpha, bigram_nonalpha, bigram_log_ratio,
         count_w1, count_w2, count_w1_and, count_w2_and,
         count_and_w1, count_and_w2)

# Load frequency log (overall_freq = total training occurrences in finetune corpus)
freqs <- read_csv(FREQ_CSV, show_col_types = FALSE) |>
  mutate(binom_alpha = paste(pmin(word1, word2), "and", pmax(word1, word2))) |>
  select(binom_alpha, overall_freq)

# Join everything
pref <- pref_raw |>
  left_join(bigrams, by = "binom_alpha") |>
  left_join(freqs,   by = "binom_alpha") |>
  mutate(
    # preference residualised against bigram log-ratio
    pref_residual = preference - bigram_log_ratio
  )

cat(sprintf("Loaded %d rows | %d models | %d unique binomials\n",
            nrow(pref), n_distinct(pref$model), n_distinct(pref$binom_alpha)))
```

---

## Plot 1a — Final checkpoint: raw preference vs bigram log-ratio (ablated models)

Do models show preferences, and are those preferences driven by bigram statistics
or something more?

```{r plot1a}
final_ablated <- pref |>
  filter(model_type == "ablated") |>
  group_by(model, model_size, binom_alpha, bigram_log_ratio) |>
  slice_max(step, n = 1) |>
  summarise(mean_pref = mean(preference), .groups = "drop")

ggplot(final_ablated, aes(x = bigram_log_ratio, y = mean_pref, colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey40") +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~model_size) +
  labs(
    title    = "Model preference vs corpus bigram log-ratio (final checkpoint, ablated)",
    subtitle = "Dotted line = perfect bigram match; slope > 1 means model amplifies corpus statistics",
    x        = "log P(and|w1)·P(w2|and) − log P(and|w2)·P(w1|and)  [bigram log-ratio]",
    y        = "Mean preference  (log P(α) − log P(non-α))",
    colour   = "Model size"
  ) +
  theme(legend.position = "bottom")
```

## Plot 1b — Final checkpoint: residual preference after accounting for bigrams

```{r plot1b}
ggplot(final_ablated |>
         mutate(pref_residual = mean_pref - bigram_log_ratio),
       aes(x = model_size, y = pref_residual, colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange",
               colour = "black", size = 0.7) +
  labs(
    title    = "Residual preference above bigram log-ratio (final checkpoint, ablated)",
    subtitle = "Each point = one binomial; black bar = mean ± 95% CI",
    x        = "Model size",
    y        = "Preference − bigram log-ratio",
    colour   = "Model size"
  ) +
  theme(legend.position = "none")
```

---

## Plot 2 — Evolution of (residualised) preference throughout training

```{r plot2, fig.height=6}
training_curve <- pref |>
  filter(model_type == "ablated", !is.na(tokens), tokens > 0) |>
  group_by(model, model_size, tokens, binom_alpha, bigram_log_ratio) |>
  summarise(mean_pref = mean(preference), .groups = "drop") |>
  mutate(pref_residual = mean_pref - bigram_log_ratio) |>
  group_by(model_size, tokens) |>
  summarise(
    mean_raw      = mean(mean_pref),
    mean_residual = mean(pref_residual),
    se_residual   = sd(pref_residual) / sqrt(n()),
    .groups = "drop"
  )

p_raw <- ggplot(training_curve,
                aes(x = tokens / 1e6, y = mean_raw, colour = model_size)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_x_continuous(labels = scales::label_number(suffix = "M")) +
  labs(title = "Raw preference over training",
       x = "Training tokens", y = "Mean preference", colour = "Size")

p_resid <- ggplot(training_curve,
                  aes(x = tokens / 1e6, y = mean_residual,
                      colour = model_size, fill = model_size)) +
  geom_ribbon(aes(ymin = mean_residual - se_residual,
                  ymax = mean_residual + se_residual), alpha = 0.15, colour = NA) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_x_continuous(labels = scales::label_number(suffix = "M")) +
  labs(title = "Residual preference (above bigram log-ratio) over training",
       x = "Training tokens", y = "Mean residual preference", colour = "Size",
       fill = "Size")

p_raw / p_resid + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

---

## Plot 3 — Ablated vs finetuned: preference shift by training frequency

For each (binomial, prefix) pair, how much does the preference change after finetuning
on the removed sentences? Does this scale with how often the binomial appeared?

```{r plot3, fig.height=6}
last_ablated <- pref |>
  filter(model_type == "ablated") |>
  group_by(model, model_size) |>
  slice_max(step, n = 1) |>
  ungroup() |>
  select(model_size, binom_alpha, prompt, pref_ablated = preference)

last_finetuned <- pref |>
  filter(model_type == "finetuned") |>
  select(model_size, binom_alpha, prompt, pref_finetuned = preference)

delta_df <- inner_join(last_ablated, last_finetuned,
                       by = c("model_size", "binom_alpha", "prompt")) |>
  mutate(pref_delta = pref_finetuned - pref_ablated) |>
  left_join(freqs, by = "binom_alpha")

delta_binom <- delta_df |>
  group_by(model_size, binom_alpha, overall_freq) |>
  summarise(mean_delta = mean(pref_delta), .groups = "drop")

if (nrow(delta_binom) == 0) {
  message("No finetuned data available yet — skipping Plot 3.")
} else {
  print(
    ggplot(delta_binom, aes(x = log(overall_freq), y = mean_delta,
                            colour = model_size)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
      geom_point(alpha = 0.5, size = 1.8) +
      geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
      facet_wrap(~model_size) +
      labs(
        title    = "Preference shift (finetuned − ablated) vs training frequency",
        subtitle = "Positive = preference for alphabetical ordering increased after finetuning",
        x        = "log(overall frequency in finetune corpus)",
        y        = "Δ preference  (finetuned − ablated)",
        colour   = "Model size"
      ) +
      theme(legend.position = "bottom")
  )
}
```

```{r plot3_regression}
# Mixed-effects model on all prompt-level observations.
# (1 | binom_alpha) accounts for correlation among 50 prompt rows per binomial.
# (1 | prompt) accounts for systematic baseline differences across prompts.
if (nrow(delta_df) == 0) {
  message("No finetuned data available yet — skipping regression.")
} else {
  delta_df |>
    group_by(model_size) |>
    group_map(~ {
      cat("\n===", .y$model_size, "===\n")
      m <- lmer(pref_delta ~ log(overall_freq) + (1 | binom_alpha) + (1 | prompt),
                data = .x, REML = FALSE)
      print(summary(m))
    })
}
```

---

## Plot 3b — Preference attenuation toward zero by training frequency

Does finetuning on equally-ordered binomials wash out model preferences more for
high-frequency items? We measure attenuation as the reduction in preference
*magnitude* after finetuning: `|pref_ablated| − |pref_finetuned|`. Positive values
mean the preference was pulled closer to zero regardless of direction.

```{r plot3b}
if (nrow(delta_df) == 0) {
  message("No finetuned data available yet — skipping Plot 3b.")
} else {
  attenuation_df <- delta_df |>
    mutate(attenuation = abs(pref_ablated) - abs(pref_finetuned)) |>
    group_by(model_size, binom_alpha, overall_freq) |>
    summarise(mean_attenuation = mean(attenuation), .groups = "drop")

  print(
    ggplot(attenuation_df, aes(x = log(overall_freq), y = mean_attenuation,
                                colour = model_size)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
      geom_point(alpha = 0.5, size = 1.8) +
      geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
      facet_wrap(~model_size) +
      labs(
        title    = "Preference attenuation toward zero vs training frequency",
        subtitle = "Positive = finetuning reduced preference magnitude; negative = strengthened it",
        x        = "log(overall frequency in finetune corpus)",
        y        = "|pref_ablated| − |pref_finetuned|",
        colour   = "Model size"
      ) +
      theme(legend.position = "bottom")
  )
}
```

```{r plot3b_regression}
# Simple OLS on per-binomial means (one averaged point per binomial)
if (exists("attenuation_df") && nrow(attenuation_df) > 0) {
  attenuation_df |>
    group_by(model_size) |>
    group_map(~ {
      cat("\n===", .y$model_size, "===\n")
      print(summary(lm(mean_attenuation ~ log(overall_freq), data = .x)))
    })
} else {
  message("No finetuned data available yet — skipping regression.")
}
```

```{r plot3b_mixed}
# Mixed-effects model on all prompt-level attenuation values.
# Note: attenuation = |pref_ablated| - |pref_finetuned|. Positive = preference
# magnitude reduced. Interpretive caveat: if pref_finetuned crosses zero (sign
# flips), attenuation is still positive even though the preference *direction*
# changed, not just its magnitude.
if (nrow(delta_df) == 0) {
  message("No finetuned data available yet — skipping mixed model.")
} else {
  delta_df |>
    mutate(attenuation = abs(pref_ablated) - abs(pref_finetuned)) |>
    group_by(model_size) |>
    group_map(~ {
      cat("\n===", .y$model_size, "(mixed model) ===\n")
      m <- lmer(attenuation ~ log(overall_freq) + (1 | binom_alpha) + (1 | prompt),
                data = .x, REML = FALSE)
      print(summary(m))
    })
}
```

---

## Plot 4 — Per-binomial preference heatmap at final checkpoint (ablated)

Which binomials develop the strongest preferences, and do they agree across model sizes?

```{r plot4, fig.height=8}
N_EXTREME <- 15

heatmap_df <- pref |>
  filter(model_type == "ablated") |>
  group_by(model, model_size, binom_alpha) |>
  slice_max(step, n = 1) |>
  summarise(mean_pref = mean(preference), .groups = "drop")

# Identify top and bottom N binomials by mean preference across model sizes
extreme_binoms <- heatmap_df |>
  group_by(binom_alpha) |>
  summarise(overall_mean = mean(mean_pref), .groups = "drop") |>
  slice_max(overall_mean, n = N_EXTREME) |>
  bind_rows(
    heatmap_df |>
      group_by(binom_alpha) |>
      summarise(overall_mean = mean(mean_pref), .groups = "drop") |>
      slice_min(overall_mean, n = N_EXTREME)
  )

heatmap_df <- heatmap_df |>
  filter(binom_alpha %in% extreme_binoms$binom_alpha) |>
  mutate(binom_alpha = fct_reorder(binom_alpha, mean_pref, .fun = mean))

ggplot(heatmap_df, aes(x = model_size, y = binom_alpha, fill = mean_pref)) +
  geom_tile() +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0, name = "Preference\n(α − non-α)") +
  geom_text(aes(label = round(mean_pref, 2)), colour = "black", size = 2.5) +
  labs(
    title    = "Per-binomial preference at final checkpoint (ablated models)",
    subtitle = glue::glue("Top and bottom {N_EXTREME} binomials by mean preference across model sizes"),
    x        = "Model size",
    y        = "Binomial (alphabetical order)"
  ) +
  theme(axis.text.y = element_text(size = 9))
```

## Plot 4b — Same binomials at 1/3 and 2/3 of training

```{r plot4b, fig.height=8}
# For each model find the checkpoints closest to 1/3 and 2/3 of max step
checkpoint_df <- pref |>
  filter(model_type == "ablated") |>
  group_by(model, model_size) |>
  mutate(
    max_step     = max(step),
    target_third = max_step / 3,
    target_two   = 2 * max_step / 3
  ) |>
  ungroup()

snap_to_fraction <- function(df, fraction_col) {
  df |>
    group_by(model, model_size, binom_alpha) |>
    slice_min(abs(step - first(!!sym(fraction_col))), n = 1) |>
    summarise(mean_pref = mean(preference), .groups = "drop")
}

heatmap_third <- snap_to_fraction(checkpoint_df, "target_third") |>
  filter(binom_alpha %in% extreme_binoms$binom_alpha) |>
  mutate(binom_alpha = factor(binom_alpha, levels = levels(heatmap_df$binom_alpha)),
         fraction = "1/3")

heatmap_two <- snap_to_fraction(checkpoint_df, "target_two") |>
  filter(binom_alpha %in% extreme_binoms$binom_alpha) |>
  mutate(binom_alpha = factor(binom_alpha, levels = levels(heatmap_df$binom_alpha)),
         fraction = "2/3")

heatmap_over_time <- bind_rows(heatmap_third, heatmap_two)

ggplot(heatmap_over_time, aes(x = model_size, y = binom_alpha, fill = mean_pref)) +
  geom_tile() +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0, name = "Preference\n(α − non-α)") +
  geom_text(aes(label = round(mean_pref, 2)), colour = "black", size = 2.5) +
  facet_wrap(~fraction, labeller = label_both) +
  labs(
    title    = "Per-binomial preference at 1/3 and 2/3 of training (ablated models)",
    subtitle = glue::glue("Same top/bottom {N_EXTREME} binomials as final checkpoint"),
    x        = "Model size",
    y        = "Binomial (alphabetical order)"
  ) +
  theme(axis.text.y = element_text(size = 9))
```

---

## Plot 5 — Distribution of preferences at final checkpoint (all models)

```{r plot5, fig.height=6}
dist_df <- pref |>
  group_by(model, model_size, model_type) |>
  slice_max(step, n = 1) |>
  ungroup() |>
  group_by(model, model_size, model_type, binom_alpha) |>
  summarise(mean_pref = mean(preference), .groups = "drop") |>
  mutate(model_type = factor(model_type, levels = c("ablated", "finetuned")))

ggplot(dist_df, aes(x = mean_pref, fill = model_type, colour = model_type)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~model_size, ncol = 1) +
  labs(
    title    = "Distribution of preferences at final checkpoint",
    subtitle = "Each curve = one model; dashed line = no preference",
    x        = "Preference  (log P(α) − log P(non-α))",
    y        = "Density",
    fill     = "Model type",
    colour   = "Model type"
  ) +
  theme(legend.position = "bottom")
```

---

## Plot 6 — Model size effect: do larger models develop stronger preferences?

```{r plot6}
size_comparison <- pref |>
  filter(model_type == "ablated") |>
  group_by(model_size, binom_alpha, bigram_log_ratio) |>
  slice_max(step, n = 1) |>
  summarise(mean_pref = mean(preference), .groups = "drop") |>
  mutate(pref_residual = mean_pref - bigram_log_ratio)

ggplot(size_comparison,
       aes(x = model_size, y = pref_residual, fill = model_size)) +
  geom_violin(alpha = 0.4, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 4, fill = "white", colour = "black") +
  labs(
    title    = "Residual preference by model size (final checkpoint, ablated)",
    subtitle = "Diamond = mean; violin quantiles at 25/50/75%",
    x        = "Model size",
    y        = "Preference − bigram log-ratio",
    fill     = "Model size"
  ) +
  theme(legend.position = "none")
```
